---
title: "Machine learning project Traditional factors"
author: "Francesco Viglietta"
date: "2023-10-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


setwd("C:/Users/Utente/Desktop/Machine Learning project")
library(readr)
library(tibble)
library(tidyverse)
library(readxl)
library(writexl)
quarterly <- read_csv("C:/Users/Utente/Desktop/Machine Learning project/QUARTERLY.csv")

factors<- quarterly[1, ]
quarterly <-quarterly[-1, ] 

quarterly<- quarterly[-(260:261), ]
#quarterly <-quarterly[, -248] 



for (i in 2:ncol(quarterly)) {
  if (quarterly[1, i] == 5) {
    column <- log(as.numeric(unlist(quarterly[, i])))
    quarterly[-1, i] <- diff(column)
  } 
  else if (quarterly[1, i] == 2) {
    column <- quarterly[, i]
    quarterly[-1, i] <- diff(as.numeric(unlist(column)), differences = 1L)
  } 
  else if (quarterly[1, i] == 3) {
    column <- quarterly[, i]
    quarterly[, i] <- diff(as.numeric(unlist(column)), differences = 2L)
  } 
  else if (quarterly[1, i] == 6) {
    column <- log(as.numeric(unlist(quarterly[, i])))
    quarterly[, i] <- c(6, NA, diff(column, differences = 2L)) 
  }
}


quarterly<- quarterly[-(1:5), ]


```















```{r}
##Defining the  Factors
real_activity_factor <- c("sasdate","USPRIV", "USGOOD", "OUTMS", "PAYEMS", "IPMANSICS", "INDPRO", "MANEMP", "HOANBS", "UNRATE", "DMANEMP")

##PERMIT INIZIA 2 ANNI DOPO; TCU DAGLI ANNI 80
forward_looking_factor <- c("sasdate","AAAFFM", "T5YFFM", "PERMIT", "BUSINVx", "HOUST", "PERMITS", "TCU", "S&P div yield", "GS10TB3Mx", "CPF3MTB3Mx")

prices_factor <- c("sasdate","CUSR0000SA0L2", "CUSR0000SAC", "DGDSRG3Q086SBEA", "PCECTPI", "CPITRNSL", "DNDGRG3Q086SBEA", "CUSR0000SA0L5", "CPIAUCSL", "WPSID61", "CPIULFSL")

fourth_factor <- c("sasdate", "CES9093000001", "CES9092000001", "USGOVT", "GFDEBTNx", "REVOLSLx", "COMPRMS", "USFIRE", "USSERV", "EXUSEU")

fifth_factor <- c("sasdate","OPHMFG","NWPIx","AWHMAN","HWIx","OPHPBS","OPHNFB","UNLPNBS","UNRATELTx","ULCMFG", "TLBSNNCBBDIx")





```




## Real activity factors  (with and without OUTMS)
```{r}
real_activity_variables <- quarterly[,real_activity_factor]
real_activity_variables<- real_activity_variables[-257,]

real_activity_variables1<- real_activity_variables[-(110:256),-4] #dialing with NA in the data
real_activity_variables2<- real_activity_variables[(110:254),]


#Real activity Factor prima del 6/1/1987 
pcaresults<- prcomp(real_activity_variables1[,-1], scale=T, center=T)
pc11loadings<-as.matrix(pcaresults$rotation[,1])
l<- scale(as.matrix(real_activity_variables1[,-1]), center = T, scale = T)
RealActivityFactor1=data.frame(Date=real_activity_variables1$sasdate, RealActivitypc_1a=c(l%*%pc11loadings))




#Real activity Factor da usare dal 6/1/1987 in poi

pcaresults1<- prcomp(real_activity_variables2[,-1], scale=T, center=T)
pc12loadings<-as.matrix(pcaresults1$rotation[,1])
l<- scale(as.matrix(real_activity_variables2[,-1]),center = T, scale = T)
RealActivityFactor2=data.frame(Date=real_activity_variables2$sasdate, RealActivitypc_1a=c(l%*%pc12loadings))





RealActivityFactor <- rbind(RealActivityFactor1, RealActivityFactor2)

```







## Five lags of real activity factor 1 and 2 
```{r}

lagged_RealActivityFactor<-RealActivityFactor%>%

  mutate(RealActivityPC1_lag1 = lag(RealActivityFactor$RealActivitypc_1a, 1),
    RealActivityPC1_lag2 = lag(RealActivityFactor$RealActivitypc_1a, 2),
    RealActivityPC1_lag3 = lag(RealActivityFactor$RealActivitypc_1a, 3),
    RealActivityPC1_lag4 = lag(RealActivityFactor$RealActivitypc_1a, 4),
    RealActivityPC1_lag5 = lag(RealActivityFactor$RealActivitypc_1a, 5))

lagged_RealActivityFactor <- lagged_RealActivityFactor[-(1:5),]





path <- "C:/Users/Utente/Desktop/Machine Learning project/Lagged_RealActivityFactor_buono.xls"
write_xlsx(lagged_RealActivityFactor, path)


```





Forward Looking factor
```{r}
forward_looking_variables <- quarterly[,forward_looking_factor]
forward_looking_variables1 <- forward_looking_variables[-(28:254),-8] #eliminating 9/1/1959-12/1/1959-3/1/1960 and the whole TCU column

forward_looking_variables1<- forward_looking_variables1[-1,] #data on which to compute the first factor 
pcaresults<- prcomp(forward_looking_variables1[,-1], scale=T, center=T)
pc11loadings<-as.matrix(pcaresults$rotation[,1])
l<- scale(as.matrix(forward_looking_variables1[,-1]), center = T, scale=T)
ForwardLookingFactor1=data.frame(Date=forward_looking_variables1$sasdate,ForwardLookingpc_1a=c(l%*%pc11loadings))

forward_looking_variables2 <- forward_looking_variables[29:254,]
forward_looking_variables2 <- forward_looking_variables2[-226,]
#forward_looking_variables2[225,9]<- 0

pcaresultsa<- prcomp(forward_looking_variables2[,-1], scale=T, center=T)
pc11aloadings<-as.matrix(pcaresultsa$rotation[,1])
l<- scale(as.matrix(forward_looking_variables2[,-1]), center=T, scale=T)
ForwardLookingFactor2=data.frame(Date=forward_looking_variables2$sasdate,ForwardLookingpc_1a=c(l%*%pc11aloadings))

ForwardLookingFactor <- rbind(ForwardLookingFactor1,ForwardLookingFactor2)


```
Here i started to compute the first factors from 	6/1/1960 without TCU, then i compute another one with TCU included as soon as it becomes available in 3/1/1967. The two are merged



## Five lags of Forward Looking factor
```{r}

lagged_ForwardLookingFactor<-ForwardLookingFactor%>%

  mutate(PC1_lag1 = lag(ForwardLookingFactor$ForwardLookingpc_1a, 1),
    PC1_lag2 = lag(ForwardLookingFactor$ForwardLookingpc_1a, 2),
    PC1_lag3 = lag(ForwardLookingFactor$ForwardLookingpc_1a, 3),
    PC1_lag4 = lag(ForwardLookingFactor$ForwardLookingpc_1a, 4),
    PC1_lag5 = lag(ForwardLookingFactor$ForwardLookingpc_1a, 5))

lagged_ForwardLookingFactor <- lagged_ForwardLookingFactor[-(1:5),]





path <- "C:/Users/Utente/Desktop/Machine Learning project/Lagged_ForwardLookingFactor_buono.xls"
write_xlsx(lagged_ForwardLookingFactor, path)





```





##  Fifth Factor
```{r}
Fifth_factor_variables <- quarterly[,fifth_factor]
columnstobeselected <- c("sasdate","NWPIx","AWHMAN","HWIx","OPHPBS","OPHNFB","UNLPNBS","UNRATELTx", "TLBSNNCBBDIx")
Fifth_factor_variables1 <- Fifth_factor_variables[1:109,columnstobeselected]
Fifth_factor_variables2 <- Fifth_factor_variables[110:254,]



pcaresults<- prcomp(Fifth_factor_variables1[,-1], scale=T, center=T)
pc11loadings<-as.matrix(pcaresults$rotation[,1])
l<- scale(as.matrix(Fifth_factor_variables1[,-1]), center = T, scale=T)
Fifth_factor1=data.frame(Date=Fifth_factor_variables1$sasdate,Fifth_factorpc_1a=c(l%*%pc11loadings))


pcaresults<- prcomp(Fifth_factor_variables2[,-1], scale=T, center=T)
pc11loadings<-as.matrix(pcaresults$rotation[,1])
l<- scale(as.matrix(Fifth_factor_variables2[,-1]), center=T, scale=T)
Fifth_factor2=data.frame(Date=Fifth_factor_variables2$sasdate,Fifth_factorpc_1a=c(l%*%pc11loadings))


Fifth_factor <- rbind(Fifth_factor1, Fifth_factor2)

```



## Five lags of fifth factor
```{r}
lagged_Fifth_factor<-Fifth_factor%>%

  mutate(Fifth_factorPC1_lag1 = lag(Fifth_factor$Fifth_factorpc_1a, 1),
    Fifth_factorPC1_lag2 = lag(Fifth_factor$Fifth_factorpc_1a, 2),
    Fifth_factorPC1_lag3 = lag(Fifth_factor$Fifth_factorpc_1a, 3),
    Fifth_factorPC1_lag4 = lag(Fifth_factor$Fifth_factorpc_1a, 4),
    Fifth_factorPC1_lag5 = lag(Fifth_factor$Fifth_factorpc_1a, 5))

lagged_Fifth_factor <- lagged_Fifth_factor[-(1:5),]

path <- "C:/Users/Utente/Desktop/Machine Learning project/Lagged_Fifth_factor_buono.xls"
write_xlsx(lagged_Fifth_factor, path)


```






##  Fourth Factor
```{r}
Fourth_factor_variables <- quarterly[, fourth_factor]

columnstobeselected1<-c("sasdate", "CES9093000001", "CES9092000001", "USGOVT", "USFIRE", "USSERV")
columnstobeselected2 <- c("sasdate", "CES9093000001", "CES9092000001", "USGOVT", "USFIRE", "USSERV","GFDEBTNx")
columnstobeselected3 <-c("sasdate", "CES9093000001", "CES9092000001", "USGOVT", "USFIRE", "USSERV","GFDEBTNx","REVOLSLx")
columnstobeselected4 <-c("sasdate", "CES9093000001", "CES9092000001", "USGOVT", "GFDEBTNx", "REVOLSLx", "COMPRMS", "USFIRE", "USSERV")
columnstobeselected5 <-c("sasdate", "CES9093000001", "CES9092000001", "USGOVT", "GFDEBTNx", "REVOLSLx", "COMPRMS", "USFIRE", "USSERV","EXUSEU")

Fourth_factor_variables1 <-Fourth_factor_variables[1:25,columnstobeselected1]
Fourth_factor_variables2 <-Fourth_factor_variables[26:33,columnstobeselected2]
Fourth_factor_variables3 <-Fourth_factor_variables[34:109,columnstobeselected3]      #dealing with NA in the data
Fourth_factor_variables4 <-Fourth_factor_variables[110:157,columnstobeselected4]
Fourth_factor_variables5 <-Fourth_factor_variables[158:254,columnstobeselected5]



pcaresults<- prcomp(Fourth_factor_variables1[,-1], scale=T, center=T)
pc11loadings<-as.matrix(pcaresults$rotation[,1])
l<- scale(as.matrix(Fourth_factor_variables1[,-1]), center=T, scale=T)
Fourth_factor1=data.frame(Date=Fourth_factor_variables1$sasdate,Fourth_factorpc_1a=c(l%*%pc11loadings))


pcaresults<- prcomp(Fourth_factor_variables2[,-1], scale=T, center=T)
pc11loadings<-as.matrix(pcaresults$rotation[,1])
l<- scale(as.matrix(Fourth_factor_variables2[,-1]),scale=T, center=T)
Fourth_factor2=data.frame(Date=Fourth_factor_variables2$sasdate,Fourth_factorpc_1a=c(l%*%pc11loadings))

pcaresults<- prcomp(Fourth_factor_variables3[,-1], scale=T, center=T)
pc11loadings<-as.matrix(pcaresults$rotation[,1])
l<- scale(as.matrix(Fourth_factor_variables3[,-1]),scale=T, center=T)
Fourth_factor3=data.frame(Date=Fourth_factor_variables3$sasdate,Fourth_factorpc_1a=c(l%*%pc11loadings))


pcaresults<- prcomp(Fourth_factor_variables4[,-1], scale=T, center=T)
pc11loadings<-as.matrix(pcaresults$rotation[,1])
l<- scale(as.matrix(Fourth_factor_variables4[,-1]), scale=T, center=T)
Fourth_factor4=data.frame(Date=Fourth_factor_variables4$sasdate,Fourth_factorpc_1a=c(l%*%pc11loadings))


pcaresults<- prcomp(Fourth_factor_variables5[,-1], scale=T, center=T)
pc11loadings<-as.matrix(pcaresults$rotation[,1])
l<- scale(as.matrix(Fourth_factor_variables5[,-1]),scale=T, center=T)
Fourth_factor5=data.frame(Date=Fourth_factor_variables5$sasdate,Fourth_factorpc_1a=c(l%*%pc11loadings))


Fourth_factor <- rbind(Fourth_factor1, Fourth_factor2,Fourth_factor3,Fourth_factor4,Fourth_factor5)

```




## Five lags of Fourth factor
```{r}
lagged_Fourth_factor<-Fourth_factor%>%

  mutate(Fourth_factorPC1_lag1 = lag(Fourth_factor$Fourth_factorpc_1a, 1),
    Fourth_factorPC1_lag2 = lag(Fourth_factor$Fourth_factor, 2),
    Fourth_factorPC1_lag3 = lag(Fourth_factor$Fourth_factorpc_1a, 3),
    Fourth_factorPC1_lag4 = lag(Fourth_factor$Fourth_factorpc_1a, 4),
    Fourth_factorPC1_lag5 = lag(Fourth_factor$Fourth_factorpc_1a, 5))

lagged_Fourth_factor <- lagged_Fourth_factor[-(1:5),]

path <- "C:/Users/Utente/Desktop/Machine Learning project/Lagged_Fourth_factor_buono.xls"
write_xlsx(lagged_Fourth_factor, path)




```

1 1-27 
2 28-35
3 36-111
4 112-159
5 160-256




Third factor
```{r}
prices_variables<-quarterly[,prices_factor]
#prices_variables<-prices_variables[-257,]

pcaresults <- prcomp(prices_variables[,-1], scale=T, center=T)
pc1loadings<- as.matrix(pcaresults$rotation[,1]) 
l <- scale(as.matrix(prices_variables[,-1]), scale=T, center=T)
price_factors <- data.frame(date=prices_variables$sasdate, Pricefactor_pc1a=c(l%*% pc1loadings))



```





Lags of the third factor
```{r}
lagged_prices_factor <- price_factors %>%
  mutate(
    Pricefactor_PC1_lag1 = lag(price_factors$Pricefactor_pc1a, 1),
    Pricefactor_PC1_lag2 = lag(price_factors$Pricefactor_pc1a, 2),
    Pricefactor_PC1_lag3 = lag(price_factors$Pricefactor_pc1a, 3),
    Pricefactor_PC1_lag4 = lag(price_factors$Pricefactor_pc1a, 4),
    Pricefactor_PC1_lag5 = lag(price_factors$Pricefactor_pc1a, 5)
  )

lagged_prices_factor <- lagged_prices_factor[-(1:5),]

file_path <- "C:/Users/Utente/Desktop/Machine Learning project/Lagged_prices_factor_buono.xlsx"

# Export the data frame to an Excel file
write_xlsx(lagged_prices_factor, path=file_path)
```




